Eagle: Making multiple-locus association mapping on a genome-wide scale routine

### 2018-08-01 paper.tex
1August refining v1

Methods for genome-wide association mapping have become progressively more sophisticated, statistically and computationally. Yet a well know limitation has persisted, the strength of association between a locus and trait is assessed only on a locus-by-locus basis. Statistically superior multiple-locus methods that detect multiple SNP-trait associations, simultaneously, have been available for some time but they have failed to become the method-of-choice. 
Here, we present Eagle for multiple-locus association mapping. Our goal was to develop a method that shifts mainstream association mapping from being single-locus to multiple-locus. We will show through the analysis of simulated and mouse data that Eagle finds true and avoids false 
associations better than completing single- and multiple-locus methods and does it faster. Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users whom may be unfamiliar with R.
\end{abstract}

@@

\iffalse

@@

We compared the computational and statistical performance of Eagle against five other multiple-locus methods for 
association mapping  These were BigRR, LMM-Lasso, glmnet, MLMM  and, r2VIM. 
BigRR, LMM-Lasso, and glmnet  implement three different regression-based regularisation methods; generalized ridge regression, 
lasso, and elastic net, respectively. Regularisation methods can handle parameter estimation problems where the number of predictors is
 far greater than the number of  samples. This is made possible through a penalty function that balances bias against variance. 
The application of regularisation methods to association mapping has appeal because it means that the SNP effects across an 
entire genome can be estimated, simultaneously.  The downside is that it is challenging, analytically, to calculate  
the statistical significance of an SNP effect. We instead calculate significance, empirically, via stability selection (see below). 
%% Permutation has been suggest as an empirical solution to this 
%% limitation \cite{shen2013novel} 
%% but we have instead opted for stability selection (see below).

Of the five multiple-locus methods, MLMM is the closest in philosophy to Eagle. It too is based on  building the 
best  linear mixed model via forward selection using the extBIC criterion. 
However, there are several differences between the two methods.
 MLMM does not make use of dimension reduction, which gives it a larger  computational footprint than Eagle. 
  It also builds its model differently to Eagle. Eagle uses a score statistic in which to identify SNPs to enter the model. 
  MLMM uses the statistical significance of a SNP when it is treated as a fixed effect in the model. 
 This necessitates fitting a separate linear mixed model for each unidentified SNP. Both are R packages but much of Eagle is written in 
 C++ code. There are significant differences in run time and memory usage (see Results).

 MLMM and Eagle are both R packages. However, much of Eagle is written in C++ code.

 and they are implemented quite differently leading to substantial 
 differences  in run time and memory usage (see Results).  Also, they differ in how a model is built. 
 Eagle uses a score statistic formed from parameter estimates of the random genetic effect in which to identify the next SNP to enter 
 the model.  MLMM uses the statistical significance of a SNP when it is treated as a fixed effect in the model. 
 This necessitates fitting a separate linear mixed model for each unidentified SNP. Fortunately,  MLMM does do this 
 efficiently.

R2VIM is quite different to the other four methods. Here, a non-parametric model-free rather than model-based statistical approach is adopted. 
R2VIM implements random forests and a new way of measuring the importance of a SNP.  
In random forests, the worth of a predictor is measured, empirically, by calculating its importance score. 
It is from these importance scores that SNP can be ordered in terms of their strength of association with a trait. 
The difficulty is in knowing what proportion of the 
highest ordered SNP are in true association. R2VIM addresses this by performing multiple parallel runs. Each run uses a different 
random seed, giving slightly different results for the random forest analyses. Within a run, a relative importance score (f) is calculated for 
each SNP. The greater f is than one, the more evidence there is that the SNP is in association with a trait.  A SNP-trait association is 
declared if f for the SNP is above some threshold value across all parallel runs. The difficulty is that the relationship between 
this threshold value and the chance of finding false positives is unknown.  The threshold value could be found via permutation but 
this would be computationally challenging for all but the smallest of association studies.

@@

We were also interested in comparing Eagle to single-locus methods to measure the difference in run times and statistical power. Even by limiting our focus to LMM-based methods, there were a number of efficient applications to choose from. 
We decided on GEMMA \cite{zhou2012genome} and FaST-LMM \cite{lippert2011fast}. These two applications have the same computational complexity \cite{zhou2012genome}, produce exact instead of approximate results, and are highly efficient computationally. They were developed at the same time, independently, but are similar theoretically. Both perform a single spectral decomposition of the relationship matrix $K$. Both use the eigenvector matrix to rotate the data. Both reformulate the log likelihood and 
restricted/residual maximum-likelihood (REML) 
log likelihood into a sum of $n$ terms that are easier to compute. The difference lies in their estimation procedure. FaST-LMM implement's the Brent's algorithm to optimise $\delta$. GEMMA instead implement's the Newton-Raphson algorithm. Newton-Raphson is more complicated in that it also requires the first and second derivatives of a function to be calculated. However, it is superior in terms of its convergence properties to the Brent algorithm.  Both applications are stand-alone computer programs, popular, and in common use.

\fi

>>>>

We present Eagle, a new method for multiple-locus association mapping. The
motivation for developing Eagle was to make multiple locus association mapping ?easy? and the method-of-choice. 
Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. doesn?t suffer from multiple testing issues c. is threshold free d. gives results that are immediately interpretable and e. has a computational footprint comparable to single-locus association mapping. 
By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and 
multiple-locus methods. We also analyse data from a published mouse study. Eagle found over two times more validated findings than the state-of-the-art 
single-locus method. 
Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users less familiar with R.

%% Old version of abstract
%%Methods for genome-wide association mapping have become progressively more sophisticated, statistically and computationally. Yet a well know limitation has persisted, the strength of association between a locus and trait is assessed only on a locus-by-locus basis. Statistically superior multiple-locus methods that detect multiple SNP-trait associations, simultaneously, have been available for some time but they have failed to become the method-of-choice. Here, we present Eagle for multiple-locus association mapping. Our goal was to develop a method that shifts mainstream association mapping from being single-locus to multiple-locus. We will show through the analysis of simulated and mouse data that Eagle finds true and avoids false associations better than completing single- and multiple-locus methods and does it faster. Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users whom may be unfamiliar with R.
\end{abstract}

@@

%%\iffalse

@@

We purposely chose  methods that differed from one another. BigRR, LMM-Lasso, and glmnet are all regression-based regularisation 
methods. However, BiggRR is based on generalised ridge regression, LMM-Lasso is based on lasso, and glmnet is based on elastic net. 
Regularisation methods make parameter estimation possible in models  where the number of predictors is far greater than the number of samples. 
They allow the strength of association between all the SNPs and trait to be measured within a single model, simultaneously. 
A limitation of these methods though is that the statistical significance of the SNP effects cannot be easily determined. 
 Due to the adaptive nature of the estimation procedures, to do this 
analytically is challenging and is an area of active research \ref{lockhart2014significance}.

The application of regularisation methods to association mapping is appealing because the strength of association between all the SNPs and 
trait can be measured simultaneously. 
 The downside is that it is challenging to calculate  
the statistical significance of a SNP effect, analytically. We instead calculate significance, empirically, via stability selection (see below). 
%% Permutation has been suggest as an empirical solution to this 
%% limitation \cite{shen2013novel} 
%% but we have instead opted for stability selection (see below).

MLMM is close in philosophy to Eagle. It too is based on  building the 
best  linear mixed model via forward selection using the extBIC criterion. 
However, there are several differences between the two methods.
 MLMM does not make use of dimension reduction, which gives it a larger  computational footprint than Eagle. 
  It also builds its model differently to Eagle. Eagle uses a score statistic in which to identify SNPs to enter the model. 
  MLMM uses the statistical significance of a SNP when it is treated as a fixed effect in the model. 
 This necessitates fitting a separate linear mixed model for each unidentified SNP. Both are R packages but much of Eagle is written in 
 C++ code. There are significant differences in run time and memory usage (see Results).

R2VIM is different to the other four methods in that it is a non-parametric model-free approach. It implements  
random forests and a new way of measuring the importance of a SNP.  
In random forests, the worth of a predictor is measured, empirically, by calculating its importance score. 
It is from these importance scores that SNPs can be ordered in terms of their strength of association with a trait. 
The difficulty is in knowing what proportion of the 
highest ordered SNPs are in true association. R2VIM addresses this by performing multiple parallel runs. Each run uses a different 
random seed, giving slightly different results for the random forest analyses. Within a run, a relative importance score ($f$) is calculated for 
each SNP. The greater $f$ is than one, the more evidence there is that the SNP is in association with the trait.  A SNP-trait association is 
declared if $f$ for the SNP is above some threshold value across all parallel runs. The difficulty is that the relationship between 
this threshold value and the chance of finding false positives is unknown.  The threshold value could be found via permutation but 
this would be computationally challenging for all but the smallest of association studies.

@@

GEMMA \cite{zhou2012genome} and FaST-LMM \cite{lippert2011fast} are of  the same computational complexity \cite{zhou2012genome}, produce exact instead of approximate results, and are highly efficient computationally.  Both perform a single spectral decomposition of the relationship matrix $K$, use  eigenvector matrix to rotate the data, 
and reformulate the (reml) log likelihood for easier computation.  They differ though in their estimation procedure. 
FaST-LMM implement's the Brent's algorithm. GEMMA instead implement's the Newton-Raphson algorithm. The  Newton-Raphson is more complicated  but has superior 
convergence properties.

===>

log likelihood into a sum of $n$ terms that are easier to compute. The difference lies in their estimation procedure. FaST-LMM implement's the Brent's algorithm to optimise $\delta$. GEMMA instead implement's the Newton-Raphson algorithm. Newton-Raphson is more complicated in that it also requires the first and second derivatives of a function to be calculated. However, it is superior in terms of its convergence properties to the Brent algorithm.  Both applications are stand-alone computer programs, popular, and in common use.

%%\fi
